Constraining gravitational wave amplitude birefringence with GWTC-3

### 2023-04-07 src/tex/ms.tex
Modify variables

\begin{abstract}
    The propagation of gravitational waves can reveal fundamental features of the structure of spacetime.
    For instance, differences in the propagation of gravitational-wave polarizations would be a smoking gun for parity violations in the gravitational sector, as expected from birefringent theories like Chern-Simons gravity.
    Here we look for evidence of amplitude birefringence in the latest LIGO-Virgo catalog (GWTC-3) through the use of birefringent templates inspired by dynamical Chern-Simons gravity.
    From 71 binary-black-hole signals, we obtain the most precise constraints on gravitational-wave amplitude birefringence yet, measuring a birefringent opacity of $\kappa =$ \variable{output/restricted_kappa_median.txt} with 90\% credibility, equivalent to an energy scale of $M_{\rm PV} \lesssim XX\, {\rm GeV}$ \mi{Thomas please provide the equivalent $M_{\rm PV}$ constraint from $|\kappa|$)}.
\end{abstract}

@@

Judging by $\mu_i/\sigma_i$, the two events that show the largest tension with $\kappa = 0$ are GW170818, for which $\mu_i / \sigma_i =$ \variable{output/GW170818_constraint.txt}, and GW200129\_065458 (henceforth GW200129), for which $\mu_i / \sigma_i =$ \variable{output/GW200129_constraint.txt}.
However, as we discuss in Sec.~\ref{sec:GW200129}, we have reason to think that the preference for $\kappa < 0$ in GW200129 might be driven by noise anomalies in the Virgo detector; with that in mind, in the next section we consider the effect of excluding this event from the joint result (we find its impact to be minimal).

@@

Both versions of the posterior support the lack of birefringence ($\mu = \sigma = 0$) within 90\% credibility; from the marginals of the result including all events, we constrain $\mu =$ \variable{output/mu_median.txt}for 90\%-credible symmetric intervals around the median, and $\sigma <$ \variable{output/sigma_median.txt}for the 90\%-credible one-sided upper limit.
However, even though $\sigma = 0$ is well supported, the $\sigma$ posterior peaks visibly away from the origin, indicating some preference for a nonzero variance.
This could be a sign of the presence of outliers in our sample.

@@

% Constraint on $\kappa$
Finally, the hierarchical result in Fig.~\ref{fig:corner_Gaussian} can be translated into an overall expectation for $\kappa_i$ under the assumption of a Gaussian distribution via Eq.~\eqref{eq:generic_posterior}.
We show the result of doing this in Fig.~\ref{fig:posterior_kappa}, where we also compare to the posterior on $\kappa$ obtained by assuming a shared value across events (same result shown as a blue band in Fig.~\ref{fig:violin_kappa}).
The hierarchical measurement leads to an expectation that $\kappa_i =$ \variable{output/generic_kappa_median.txt}, whereas the shared-$\kappa$ measurement implies $\kappa =$ \variable{output/restricted_kappa_median.txt}, both symmetric 90\%-credible intervals around the median.

@@

Our primary constraint on birefringence is given by the combined constraint on the birefringent opacity from GWTC-3, $\kappa =$ \variable{output/restricted_kappa_median.txt} with 90\% credibility, obtained in Sec.~\ref{sec:results:gwtc} (blue band in Fig.~\ref{fig:violin_kappa}).
We can compare this to previous measurements of amplitude birefringence in the literature.

\subsubsection{Okounkova et al.}
\citet{Okounkova_2022} produced a constraint on frequency-independent \ac{GW} amplitude birefringence from the distribution of measured inclinations of \acp{BBH} in GWTC-2.
As opposed to our Eq.~\eqref{eq:waveform_modification}, this reference parameterized the effect of birefringence as
\begin{equation}
\label{eq:freqindep}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm \tilde{\kappa}\, d_C\right)\,,
\end{equation}
where $\tilde{\kappa}$ is a frequency-independent opacity (simply denoted ``$\kappa$'' in Ref.~\cite{Okounkova_2022}).
Assuming a fiducial redshift of $z=0.3$ for all events in GWTC-2, \citet{Okounkova_2022} obtained a constraint of $|\tilde{\kappa}| \lesssim 0.74\, {\rm Gpc}^{-1}$ at 68\% credibility.
Taking a reference frequency of $f=100\, \mathrm{Hz}$, which is representative of LIGO-Virgo events, and adopting the same fiducial distance $d_C(z=0.3)$, we may compare that constraint to ours by noting that
\begin{equation}
    \left[ \kappa = \tilde{\kappa}\left({1\, \mathrm{Gpc}}\right) \right]_{\raisebox{1ex}{$\begin{subarray}{l} z=0.3 \\ f=100\,{\rm Hz} \end{subarray}$}}
\end{equation}
per Eq.~\eqref{eq:waveform_modification}.
In terms of this fiducial quantity, the constraint given by \citet{Okounkova_2022} can be expressed by $|{\kappa}| \lesssim 0.74$  at 68\% credibility.
On the other hand, our constraint from Sec.~\ref{sec:results:gwtc} is $|\kappa| <$ \variable{output/restricted_absolute_kappa_68.txt} at 68\% credibility.
This is a factor of \variable{output/improvement_Okounkova.txt} more stringent than Ref.~\cite{Okounkova_2022}, besides arising from a less simplified model.

@@

\citet{Wang_2021} performed \ac{PE} on GWTC-1 events with a frequency-dependent model of birefringence resembling ours.
In contrast to our Eq.~\eqref{eq:waveform_modification}, this reference parameterized the birefringent waveform in terms of some amplitude and phase modifications to the linear polarizations ($\delta h$ and $\delta \Psi$ respectively), such that
\begin{equation}
    h_{+/\times}^{\rm BR}(f) = h_{+/\times}^{\rm GR}(f)\mp h_{\times/+}^{\rm GR}(f)(i\delta h-\delta\Psi)\,,
\end{equation}
with the plus (minus) sign for the cross (plus) polarization.
In terms of circular polarizations, this is equivalent to 
\begin{align}
h^{\rm BR}_{L/R}(f) &= h^{\rm GR}_{L/R}(f) \left(1 \mp \delta h \mp i \delta \Psi \right) \nonumber\\
&\approx h^{\rm GR}_{L/R}(f) \exp(\mp \delta h \mp i \delta \Psi) \, ,
\end{align}
assuming a small $\delta $ and $\delta \Psi \ll 1$.
To consider only amplitude birefringence, as in this work, we must compare to the result in Ref.~\citep{Wang_2021} that set $\delta \Psi = 0$ and parameterized $\delta h = \pi f z / M_{\rm PV}$, where $M_{\rm PV}$ is the energy scale of the birefringent (parity-violating) correction in units of frequency.%
\footnote{Concretely, \citet{Wang_2021} write $\delta h = - A_\nu \pi f$ with $A_\nu = M_{\rm PV}^{-1} \left[\alpha_\nu (z=0) - \alpha_\nu(z) \left(1+z\right)\right]$, for $\alpha_\nu$ some function of redshift encoding the evolution of a birefringence-mediating field; to produce their constraint, however, they further set $\alpha_\nu(z) = 1$, yielding $A_\nu = -z / M_{\rm PV}$ and hence $\delta h = \pi f z / M_{\rm PV}$.}
Comparing to our Eq.~\eqref{eq:waveform_modification} and approximating $z \approx d_C H_0/c$ via the Hubble constant, $H_0$, this means
\begin{equation}
    |\kappa| \simeq \frac{\pi H_0}{c}\left(1\,\mathrm{Gpc}\right)\left(100\,\mathrm{Hz}\right)M_{\rm PV}^{-1}\,.
\end{equation}
\citet{Wang_2021} quote a constraint of $M_{\rm PV} > 10^{-22}\, {\rm GeV}$ at 90\% credibility, which translates into $|\kappa| \lesssim$ \variable{output/kappa_Wang.txt}.
We obtained a tighter 90\% upper limit of $|\kappa| <$ \variable{output/restricted_absolute_kappa_90.txt}.

@@

We found no evidence of amplitude birefringence in the GWTC-3 data, and constrained $\kappa =$ \variable{output/restricted_kappa_median.txt}with 90\% credibility when treating $\kappa$ as a global quantity shared by all GWTC-3 events (Sec.~\ref{sec:results:gwtc}).
As an additional check, we implemented a hierarchical analysis that allowed for each event to probe different effective values of $\kappa$, as might be the case if birefringence is mediated by a field that is nonuniform over the angular separations between detected sources (Sec.~\ref{sec:results:hier}).
The result of that analysis was consistent with a vanishing $\kappa$ for all events in our catalog within 90\% credibility, but hinted at some possible variance in the population.
Future catalogs with more events will shed light on this feature.

>>>>

\begin{abstract}
    The propagation of gravitational waves can reveal fundamental features of the structure of spacetime.
    For instance, differences in the propagation of gravitational-wave polarizations would be a smoking gun for parity violations in the gravitational sector, as expected from birefringent theories like Chern-Simons gravity.
    Here we look for evidence of amplitude birefringence in the latest LIGO-Virgo catalog (GWTC-3) through the use of birefringent templates inspired by dynamical Chern-Simons gravity.
    From 71 binary-black-hole signals, we obtain the most precise constraints on gravitational-wave amplitude birefringence yet, measuring a birefringent opacity of \variable{output/restricted_kappa_median.txt}with 90\% credibility, equivalent to an energy scale of \variable{output/M_PV_constraint.txt}.
\end{abstract}

@@

Judging by $\mu_i/\sigma_i$, the two events that show the largest tension with $\kappa = 0$ are GW170818, for which \variable{output/GW170818_constraint.txt}, and GW200129\_065458 (henceforth GW200129), for which \variable{output/GW200129_constraint.txt}.
However, as we discuss in Sec.~\ref{sec:GW200129}, we have reason to think that the preference for $\kappa < 0$ in GW200129 might be driven by noise anomalies in the Virgo detector; with that in mind, in the next section we consider the effect of excluding this event from the joint result (we find its impact to be minimal).

@@

Both versions of the posterior support the lack of birefringence ($\mu = \sigma = 0$) within 90\% credibility; from the marginals of the result including all events, we constrain \variable{output/mu_median.txt}for 90\%-credible symmetric intervals around the median, and \variable{output/sigma_median.txt}for the 90\%-credible one-sided upper limit.
However, even though $\sigma = 0$ is well supported, the $\sigma$ posterior peaks visibly away from the origin, indicating some preference for a nonzero variance.
This could be a sign of the presence of outliers in our sample.

@@

% Constraint on $\kappa$
Finally, the hierarchical result in Fig.~\ref{fig:corner_Gaussian} can be translated into an overall expectation for $\kappa_i$ under the assumption of a Gaussian distribution via Eq.~\eqref{eq:generic_posterior}.
We show the result of doing this in Fig.~\ref{fig:posterior_kappa}, where we also compare to the posterior on $\kappa$ obtained by assuming a shared value across events (same result shown as a blue band in Fig.~\ref{fig:violin_kappa}).
The hierarchical measurement leads to an expectation that \variable{output/generic_kappa_median.txt}, whereas the shared-$\kappa$ measurement implies \variable{output/restricted_kappa_median.txt}, both symmetric 90\%-credible intervals around the median.

@@

Our primary constraint on birefringence is given by the combined constraint on the birefringent opacity from GWTC-3, \variable{output/restricted_kappa_median.txt}with 90\% credibility, obtained in Sec.~\ref{sec:results:gwtc} (blue band in Fig.~\ref{fig:violin_kappa}).
We can compare this to previous measurements of amplitude birefringence in the literature.

\subsubsection{Okounkova et al.}
\citet{Okounkova_2022} produced a constraint on frequency-independent \ac{GW} amplitude birefringence from the distribution of measured inclinations of \acp{BBH} in GWTC-2.
As opposed to our Eq.~\eqref{eq:waveform_modification}, this reference parameterized the effect of birefringence as
\begin{equation}
\label{eq:freqindep}
    h_{L/R}^{\mathrm{br}}(f) =
    h_{L/R}^{\mathrm{GR}}(f) \times
    \exp\left(\pm \tilde{\kappa}\, d_C\right)\,,
\end{equation}
where $\tilde{\kappa}$ is a frequency-independent opacity (simply denoted ``$\kappa$'' in Ref.~\cite{Okounkova_2022}).
Assuming a fiducial redshift of $z=0.3$ for all events in GWTC-2, \citet{Okounkova_2022} obtained a constraint of $|\tilde{\kappa}| \lesssim 0.74\, {\rm Gpc}^{-1}$ at 68\% credibility.
Taking a reference frequency of $f=100\, \mathrm{Hz}$, which is representative of LIGO-Virgo events, and adopting the same fiducial distance $d_C(z=0.3)$, we may compare that constraint to ours by noting that
\begin{equation}
    \left[ \kappa = \tilde{\kappa}\left({1\, \mathrm{Gpc}}\right) \right]_{\raisebox{1ex}{$\begin{subarray}{l} z=0.3 \\ f=100\,{\rm Hz} \end{subarray}$}}
\end{equation}
per Eq.~\eqref{eq:waveform_modification}.
In terms of this fiducial quantity, the constraint given by \citet{Okounkova_2022} can be expressed by $|{\kappa}| \lesssim 0.74$  at 68\% credibility.
On the other hand, our constraint from Sec.~\ref{sec:results:gwtc} is \variable{output/restricted_absolute_kappa_68.txt}at 68\% credibility.
This is a factor of \variable{output/improvement_Okounkova.txt}more stringent than Ref.~\cite{Okounkova_2022}, besides arising from a less simplified model.

@@

\citet{Wang_2021} performed \ac{PE} on GWTC-1 events with a frequency-dependent model of birefringence resembling ours.
In contrast to our Eq.~\eqref{eq:waveform_modification}, this reference parameterized the birefringent waveform in terms of some amplitude and phase modifications to the linear polarizations ($\delta h$ and $\delta \Psi$ respectively), such that
\begin{equation}
    h_{+/\times}^{\rm BR}(f) = h_{+/\times}^{\rm GR}(f)\mp h_{\times/+}^{\rm GR}(f)(i\delta h-\delta\Psi)\,,
\end{equation}
with the plus (minus) sign for the cross (plus) polarization.
In terms of circular polarizations, this is equivalent to 
\begin{align}
h^{\rm BR}_{L/R}(f) &= h^{\rm GR}_{L/R}(f) \left(1 \mp \delta h \mp i \delta \Psi \right) \nonumber\\
&\approx h^{\rm GR}_{L/R}(f) \exp(\mp \delta h \mp i \delta \Psi) \, ,
\end{align}
assuming a small $\delta $ and $\delta \Psi \ll 1$.
To consider only amplitude birefringence, as in this work, we must compare to the result in Ref.~\citep{Wang_2021} that set $\delta \Psi = 0$ and parameterized $\delta h = \pi f z / M_{\rm PV}$, where $M_{\rm PV}$ is the energy scale of the birefringent (parity-violating) correction in units of frequency.%
\footnote{Concretely, \citet{Wang_2021} write $\delta h = - A_\nu \pi f$ with $A_\nu = M_{\rm PV}^{-1} \left[\alpha_\nu (z=0) - \alpha_\nu(z) \left(1+z\right)\right]$, for $\alpha_\nu$ some function of redshift encoding the evolution of a birefringence-mediating field; to produce their constraint, however, they further set $\alpha_\nu(z) = 1$, yielding $A_\nu = -z / M_{\rm PV}$ and hence $\delta h = \pi f z / M_{\rm PV}$.}
Comparing to our Eq.~\eqref{eq:waveform_modification} and approximating $z \approx d_C H_0/c$ via the Hubble constant, $H_0$, this means
\begin{equation}
    |\kappa| \simeq \frac{\pi H_0}{c}\left(1\,\mathrm{Gpc}\right)\left(100\,\mathrm{Hz}\right)M_{\rm PV}^{-1}\,.
\end{equation}
\citet{Wang_2021} quote a constraint of $M_{\rm PV} > 10^{-22}\, {\rm GeV}$ at 90\% credibility, which translates into \variable{output/kappa_Wang.txt}.
We obtained a tighter 90\% upper limit of \variable{output/restricted_absolute_kappa_90.txt}.

@@

We found no evidence of amplitude birefringence in the GWTC-3 data, and constrained \variable{output/restricted_kappa_median.txt}with 90\% credibility when treating $\kappa$ as a global quantity shared by all GWTC-3 events (Sec.~\ref{sec:results:gwtc}).
As an additional check, we implemented a hierarchical analysis that allowed for each event to probe different effective values of $\kappa$, as might be the case if birefringence is mediated by a field that is nonuniform over the angular separations between detected sources (Sec.~\ref{sec:results:hier}).
The result of that analysis was consistent with a vanishing $\kappa$ for all events in our catalog within 90\% credibility, but hinted at some possible variance in the population.
Future catalogs with more events will shed light on this feature.
